Parallel Adaptive Importance Sampling

### 2018-02-16 response_JUQ.tex

\usepackage[colorlinks]{hyperref}

\newcommand{\comment}[2]{\vspace{0.6cm}{\bf Comment:} {\it #1.}

\vspace{0.3cm}{\bf Answer:} #2}

\begin{document}
\title{Response to Referee Comments: Parallel Adaptive Importance Sampling}
\maketitle
First of all, we would like to thank the referees and the associate editor for
their time in reading and commenting on our manuscript. In this document
we will address each of the points raised, and describe any relevant
changes that have been made to the paper. We have made some substantial changes, not least with the inclusion of a new ``large scale'' example where the likelihood requires a solve of a differential equation. We demonstrate in this example that PAIS is 2 orders of magnitude faster, in terms of CPU time, than the equivalent standard MCMC.

@@

\comment{For mixture problems, if the mixture has to many 
elements (or the number of elements is unknown), then this scheme will have 
difficulties.}{This is true, as would state-of-the-art Riemannian manifold MALA, HMC and a whole range of other methods. I would be happy for the referee to indicate a method which would comfortably deal with a mixture problem with a large number of possible known/unknown components. The example was chosen to demonstrate the remarkable performance of the algorithm (with respect to standard MCMC methods) for multimodal posteriors in low dimensions, and we believe it has achieved that. We have added a paragraph to the end of the example to discuss the point that the referee makes.}%done

\comment{Also, it is not clear how it will behave in real life inverse 
problems where the likelihood is determined by complex physical systems that 
involve, for example, ODE's or PDE's. If the likelihood is expensive to 
evaluate, the tunning of the scaling parameters for the PAIS-RW might be a 
concern when the number of chains is big.}{The tuning of the scaling parameters is actually very fast, often converging within a few percentage points of the optimum within the order of $10^2$ iterations. We have included some plots and discussion to the end of Section 4.2 to demonstrate this point, and we thank the referee for highlighting this issue. This is quicker than one might expect for an adaptive serial algorithm.}%done

\comment{An real data example for the PAIS algorithm is much needed. It will be 
interesting to see how PAIS is compared not only against a naive 
parallelization, but also compared against a state of the art MCMC algorithm 
or/and a efficient ad hoc MCMC algorithm designed for such example.}{Unfortunately we do not have the resources to apply our method to a full-scale real data example. However, following additional comments from the editor, we have included a new example where the likelihood requires the solution of an ODE. In Section 7.5 we include an example of data assimilation for the Lorenz '63 equations, which is a set of equations which is commonly used to test methods for inference in the applied mathematics community.

Expensive likelihoods actually make this method more applicable, since they make the extra overheads of the resampling step relatively less important. This is precisely what the method is designed for, and is the very reason for us to compare the algorithms by looking at error as a function of the number of samples/likelihood evaluations. This is seen even for a simple 3D ODE as we see in our example in Section 7.5, where PAIS reaches a relative error in the mean of $10^{-5}$ in a time which is 2 orders of magnitude quicker in CPU time than for trivially parallelised RWMH chains.}%done

@@

\comment{There aren't details about the architecture used to run the M chains in 
parallel. This is relevant since there are several options such as: multi-core 
and multi-processor computers, GPU cards, clusters, cloud computing, etc. 
Comments about the advantages and disadvantages of the architecture 
are also welcome. The same goes for details regarding the programming language 
used.}{We thank the referee for this comment, as we should have included these details in the manuscript. The method was implemented in C++, serially, and was run on a single core of a Dell server with four 8-core 3.3GHz CPUs, 64Gb memory, and these details have been added in Section 7.

We are not computer scientists, and as such, we are not in a position to comment on specific architectures, especially as the optimal architecture would be problem specific. The optimal architecture would likely depend on the relative costs of communication of the states across the processors, the cost of the likelihood evaluation, and the cost of the resampler for the chosen ensemble size. We have added a discussion to Section 3, following the presentation of the PAIS algorithm.}%done

@@

\comment{CPU time must be reported to compare the gain of each algorithm in 
order to have a fair comparison. Authors claim that there is a significant 
gain in terms of the number of iterations needed to achieve a desire 
tolerance; examples in sections 7.2 and 7.3 show that PAIS saves up to 90\% of 
the iterations. However, if the CPU time needed to run 10 PAIS iterations is 
ten times the time needed to run 100 RWMH iterations, then the gain is not as 
good as presented. PAIS iterations are more time consuming than RWMH iterations 
due to the re-sampling step. The authors mentioned that the re-sampling step is 
computationally expensive and that is why they proposed the AMR re-sampling 
method. CPU time is of particular importance in example 7.4 since the number of 
parallel chains grows from 50 to 500.}{The referee may have missed Figure 5.1 in which we show the computational cost, in time, of the resamplers that we use in the paper, both the optimal transport resampler, and the AMR. As mentioned before in response to another of the referee's comments, we had neglected to share the specs of the machine that these times were recorded on, and this has now been added. % There is an additional cost in the algorithm over and above the resampler, which is the computation of the denominator in the weights, as pointed out by the second referee. If we have $M$ members of our ensemble, then the cost of this denominator across the whole ensemble is $M^2$, but it can be easily parallelised. We include some timings of the cost of this aspect of the algorithm in Section ****, and a discussion of the overall additional cost of the approach. 
% Coupled with the timings presented for the resampler, we believe that this now gives the reader a clear idea of the additional overhead costs which are incurred when using our method as opposed to a standard MCMC method. We thank the referee for their comments on this.

The reason for analysing the error as a function of the number of the likelihood evaluations, is that this method is intended for problems with expensive likelihoods, for which these overhead costs would be dwarfed. We have added a discussion on this to the beginning of Section 7.

Additionally to this, in Section 7.5, as detailed above, we have compared the CPU time that it takes the PAIS to reach a relative error in the mean of the posterior of $10^{-5}$ against that for trivially parallelised (non-communicative) RWMH. Even for a simple example such as this, PAIS outperforms standard parallelised MCMC by two orders of magnitude.}%done

@@

\comment{4) The references should be completed with other important contributions (considering also the same kind of weights, in some cases):

- O.Cappe, R.Douc, A.Guillin, J.M.Marin, C.P.Robert, Adaptive importance sampling in general mixture classes, Statistics and Computing 18 (2008) 447-459. 
- R. Douc, A. Guillin, J. M. Marin, C. P. Robert, Convergence of adaptive mixtures of importance sampling schemes, Annals of Statistics 35 (2007) 420-448. 
- R. Douc, A. Guillin, J. M. Marin, C. P. Robert, Minimum variance importance sampling via population Monte Carlo, ESAIM: Probability and Statistics 11 (2007) 427-447. 
- L. Martino, V. Elvira, D. Luengo, J. Corander, Layered Adaptive Importance Sampling, Statistics and Computing, Vol. 27 (3), pp. 599-623, 2017.}{We thank the referee for thier suggestions for other highly relevant references.} %done

\section*{Associate Editor comments}
\comment{A very similar version of the authors’ PAIS algorithm has been studied before, as explained by the second referee. Moreover, I could argue that the PAIS algorithm is a particular case since weights are considered deter- ministic. This is a serious issue in the paper, let alone that the authors are unaware of the series of papers mentioned by referee 2, that are closely related to their work.
This should be explained in detail and the paper original perspective and contributions be further highlighted.}{Following suggestions by the second referee and the associate editor, we have added a significant new section to the introduction to highlight the other ensemble importance schemes already in the literature. We have also highlighted further the contribution made by this manuscript in the introduction and discussion sections.}%done

\comment{An other serious concern is why not use PMC-MCMC for the updating scheme in Algorithm 1. That would be as costly (since it also involves evaluating the posterior) but the yis will have the correct distribution.}{This comment is not clear to us - what is the ``updating scheme'' in Algorithm 1? Also, the (weighted) $y_i$ are distributed according to the posterior, as discussed in Section 6. We'd be happy to address this point if the AE is able to clarify it for us.}%nothing to do

### 2018-02-16 PAIS.tex
Simon final additions

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and
  complexity, the need for efficient methods which can exploit the
  parallel architectures which are prevalent in high performance
  computing increases. In this paper, we aim to develop a framework
  for scalable parallel sampling algorithms. At each iteration, an
  importance sampling proposal distribution is formed using the
  an ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this parallel
  adaptive importance sampling (PAIS) method outperforms naive
  parallelisation of serial MCMC methods for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, approximate multinomial resampling (AMR), which while not
  as accurate as other schemes is substantially less costly for large
  ensemble sizes, which can then be used in conjunction with PAIS for
  complex problems. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

To produce one sample from the posterior distribution for this example using the PAIS algorithm takes on average $5.5\times 10^{-4}$ seconds, while a sample can be produced in $1.7\times 10^{-5}$ seconds using the MH algorithm. However; the quality of the sample produced by PAIS is much higher. To reach a tolerance of $5\times 10^{-5}$, it takes the PAIS algorithm roughly 16.5 seconds and the MH algorithm roughly 1020 seconds.

@@

We have demonstrated that the framework that we have considered, with
the use of state-of-the-art optimal transport-based resampling, can
reduce the number of likelihood evaluations required to characterise
complex posterior distributions in low dimensions to a given
degree. We have also introduced a greedy approximation to this
resampler, which drastically reduces the cost, at the loss of some
accuracy, which can be mitigated with the use of larger
ensembles. Lastly, we have detailed how scaling parameters in the MCMC
proposals that are used within the mixture distribution can be quickly
and efficiently
tuned in an automated way. However, tuning the variances and
covariances of the mixture proposal components
globally is likely to be of limited use for multimodal problems where
the modes have very different covariances, or where there is curved
ridges in the density. There are a variety of possible solutions to
this problem that would be worthy of future consideration, not least
the potential to use transport maps\cite{el2012bayesian,parno2014transport} to map Gaussian mixtures to
highly complex non-Gaussian approximations of the posterior. Such a
map could encode local covariance information, and lead to accelerated
and stable PAIS sampling. This is an interesting area for future study.

@@

MCMC
RWMH
MALA
AIS
PMC
PAIS
ETPF
AMR
PAIS-RW
PAIS-MALA
APAIS-RW
APAIS-MALA

>>>>

\author{Colin Cotter\thanks{Department of Mathematics, Imperial
    College, London, UK} \and Simon Cotter\thanks{School of
    Mathematics, University of Manchester, Manchester, UK. e:
    simon.cotter@manchester.ac.uk. SLC is grateful for EPSRC First
    grant award EP/L023393/1} \and Paul Russell\thanks{School of
    Mathematics, University of Manchester, Manchester, UK}}
\title{Parallel Adaptive Importance Sampling}
\begin{document}
\maketitle
\begin{abstract}
  Markov chain Monte Carlo methods are a powerful and commonly used
  family of numerical methods for sampling from complex probability
  distributions. As applications of these methods increase in size and
  complexity, the need for efficient methods which can exploit the
  parallel architectures which are prevalent in high performance
  computing increases. In this paper, we aim to develop a framework
  for scalable parallel sampling algorithms. At each iteration, an
  importance sampling proposal distribution is formed using the
  an ensemble of particles. A stratified sample is taken from this
  distribution and weighted under the posterior, a state-of-the-art
  resampling method is then used to create an evenly weighted sample
  ready for the next iteration. We demonstrate that this parallel
  adaptive importance sampling (PAIS) method outperforms naive
  parallelisation of serial MCMC methods for low dimensional problems,
  and in fact shows
  better than linear improvements in convergence rates with respect to
  the number of ensemble members. We also introduce a new resampling
  strategy, approximate multinomial resampling (AMR), which while not
  as accurate as other schemes is substantially less costly for large
  ensemble sizes, which can then be used in conjunction with PAIS for
  complex problems. In particular, we demonstrate this methodology's
  superior sampling for multimodal problems, such as those arising
  from inference for mixture models, and for problems with expensive
  likelihoods requiring the solution of a differential equation.
\end{abstract}
\begin{keywords}MCMC, parallel, importance sampling, Bayesian, inverse problems.
\end{keywords}
\section{Introduction}
Having first been developed in the early 1970s\cite{hastings1970monte}, Markov chain Monte Carlo (MCMC) methods have been of increasing
importance and interest in the last 20 years or so. They allow us to
sample from complex probability distributions which we would not be
able to sample from directly. In particular, these methods have
revolutionised the way in which inverse problems can be tackled,
allowing full posterior sampling when using a Bayesian framework.

@@

To produce one sample from the posterior distribution for this example
using the PAIS algorithm takes on average $5.5\times 10^{-4}$ seconds,
while a sample can be produced in $1.7\times 10^{-5}$ seconds using
the MH algorithm. However; the quality of the sample produced by PAIS
is much higher. To reach a tolerance of $5\times 10^{-5}$, it takes
the PAIS algorithm roughly 16.5 seconds and the MH algorithm roughly
1020 seconds. Although we do not claim that either codes are optimally
efficient, this give an indication of the saving that can be achieved
through this type of methodology.

@@

We have demonstrated that the framework that we have considered, with
the use of state-of-the-art optimal transport-based resampling, can
reduce the number of likelihood evaluations required to characterise
complex posterior distributions in low dimensions to a given
degree. We have also introduced a greedy approximation to this
resampler, which drastically reduces the cost, at the loss of some
accuracy, which can be mitigated with the use of larger
ensembles. We have detailed how scaling parameters in the MCMC
proposals that are used within the mixture distribution can be quickly
and efficiently
tuned in an automated way. Lastly we have demonstrated that when the
likelihood is expensive, for instance because it involves the
numerical approximation of the solution of a differential equation, the savings made in terms of the number
iterations required in order reach a particular degree of accuracy
outweight the additional overhead costs that are incurred
over-and-above standard MCMC.

However, tuning the variances and
covariances of the mixture proposal components
globally is likely to be of limited use for multimodal problems where
the modes have very different covariances, or where there is curved
ridges in the density. There are a variety of possible solutions to
this problem that would be worthy of future consideration, not least
the potential to use transport maps\cite{el2012bayesian,parno2014transport} to map Gaussian mixtures to
highly complex non-Gaussian approximations of the posterior. Such a
map could encode local covariance information, and lead to accelerated
and stable PAIS sampling. This is an interesting area for future study.

@@

\begin{table}[h!]
\centering
\begin{tabular}{|l|l|}
\hline
Acronym & Full name                              \\ \hline
MCMC    & Markov chain Monte Carlo               \\
RWMH    & Random walk Metropolis-Hastings        \\
MALA    & Metropolis-adjusted Langevin algorithm \\
AIS     & Adaptive importance sampling           \\
PMC     & Population Monte Carlo                 \\
PAIS    & Parallel adaptive importance sampling  \\
ETPF    & Ensemble transport particle filter     \\
AMR     & Approximate multinomial resampler      \\
PAIS-X  & PAIS with X kernels                    \\
APAIS-X & Adaptive PAIS with X kernels          \\ \hline
\end{tabular}
\end{table}
